Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-07-10 Manuscript/manuscript.tex
Discussion edits
Initial discussion edits. Removed first paragraph about effect of
longevity.  Changes to network lit discussion.

The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.  Using a model parameterized for chronic wasting disease in mule deer, \cite{Almberg2011} find that increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the importance of reservoir transmission increasing through time.  Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.  \textcolor{red}{Other examples.}  Look into George et al (2013).

By modeling the spatial spread of such an environmentally-transmitted pathogen, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).  The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed as a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).  However, as the alternative parameterizations show, the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.  In the focal model, the dynamics of high variance metapopulations were driven largely by the presence of high quality patches, whereas in the models with the lattice structure and with equal extinction rates across all patches, high and low quality patches were more balanced in their effects, thus canceling each other out as variance increased.

\textcolor{red}{Emphasize own results relative to the contact network literature.  Should try to avoid the "main model results suggest X, but sensitivity analysis models suggest Y" format and instead try to present a more complete picture/story.  In the primary model, the increase in pathogen spread from high variance metapopulations likely results from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.  The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.}

The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.  Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).

%However, the net effect of low and high quality patches on metapopulation occupancy depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.  When there is little pathogen spread (e.g. low longevity, or weak direct infection), high quality patch distributions support greater total occupancy than low quality patch distributions
%
%At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.  In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.  Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e. high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.  In these cases, the role of patch quality lies primarily in its effect on extinction rates.  Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infectious occupancy, resulting in an increase in host extinction events at the highest longevities.

>>>>

By modeling the spatial spread of an environmentally-transmitted pathogen, these simulations suggest that highly heterogeneous metapopulations are more prone to widespread pathogen spread than more homogeneous metapopulations (Fig~\ref{poutcome}).  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).  The mechanisms underlying this behavior can be explored by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed as a network, with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

However, by including environmental transmission, our models suggest that the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.  In the baseline model, the increase in pathogen spread in high variance metapopulations results primarily from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.  The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.

The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches on pathogen spread and metapopulation stability.  Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).  \textcolor{red}{Maybe also note that natural mortality on low quality patches is high enough that there's not enough time for susceptibles to to get infected.}
